---
title: "Dynamic Data Science Applications in Optimal Profit Algorithmic Trading"
output:
  html_document: default
  pdf_document: default
fontsize: 12pt  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The file contains the source code of the data applications of the trading strategies presented in *Dynamic Data Science Applications in Optimal Profit Algorithmic Trading*.

# Load required packages
```{r, message=FALSE, warning=FALSE}
pkg_list = c('quantmod', 'TTR', 'zoo', 'tseries', 'fGarch','PEIP','gdata',
             'gridExtra','tidyverse', 'aTSA', 'dygraphs', 'urca')
# Function to install required packages if needed
for (pkg in pkg_list)
{
  # Loading the library.
  if (!library(pkg, logical.return=TRUE, character.only=TRUE))
    {
         # If the library cannot be loaded, install first and then load.
        install.packages(pkg)
        library(pkg, character.only=TRUE)
  }
}
```
# Load the required stocks
```{r, warning=FALSE, message=FALSE}
start.date = '2017-2-1' # starting date of stock
end.date = '2019-3-17' # ending date of stock
# Download the selected stocks from Yahoo finance
getSymbols(c('SPY','ADBE','EBAY','MSFT','IBM','GLD','GDX','EWA','EWC','IGE'), 
                      src = "yahoo", from = start.date, to = end.date)
stocks <-merge(SPY = SPY[, "SPY.Adjusted"], ADBE = ADBE[, "ADBE.Adjusted"], 
               EBAY= EBAY[, "EBAY.Adjusted"], MSFT = MSFT[, "MSFT.Adjusted"], 
               IBM = IBM[, "IBM.Adjusted"], GLD = GLD[, "GLD.Adjusted"], 
               GDX = GDX[, "GDX.Adjusted"], EWA = EWA[, "EWA.Adjusted"], 
               EWC = EWC[, "EWC.Adjusted"], IGE = IGE[, "IGE.Adjusted"])
head(stocks)
```
## Function to select cointegrated pairs 
```{r}
find_cointegrated_pairs <- function (data){
  n <- ncol (data)
  pvalue_matrix <- matrix(0, nrow=n, ncol=n)
  pairs <- list()
  m <- 1
  for (i in 1:n){
    for (j in 1:n){
      if(i>=j) { 
        next;
      } else{
      S1 <- data[, i]
      S2 <- data[, j]
      result <- coint.test (as.numeric(S1), as.numeric(S2), output = FALSE)
      pvalue_matrix [i, j] <- result[,3][[1]]
      if (result[,3][[1]] < 0.05){
        pairs [[m]] <- c (i, j)
        m <- m+1}
      }
    }
  }
  newlist <- list(pvalue_matrix, pairs)
  return (newlist)
}
```
## Calculate p-value to identify cointegrated pairs
```{r}
pvalue<-find_cointegrated_pairs(stocks)
#round(pvalue[[1]],3)
```
## Plot the stock prices and test for multiple cointegration 
```{r}
assets <- c("EWA", "EWC") # selected two assets
pair.stock <- merge(stocks[, 8], stocks[, 9], join="inner")
colnames(pair.stock) <- assets
# Plot the assets 
plot(pair.stock, legend.loc=1)
# Test of multiple cointegration
jotest=ca.jo(pair.stock, type="trace", K=2, ecdet="none", spec="longrun")
summary(jotest)
```

# Implementation of Non-Gaussian filter algorithm for two stocks
```{r}
# select pairwise cointegrated stocks
x <- pair.stock[, 1]
y <- pair.stock[, 2]
x$intercept <- rep(1, nrow(x)) # create intercept
var_e <- 0.0001 # innovation covariance of observation
sigma_v <- var_e/(1-var_e)*diag(2) #covariance matrix of state
Ve <- 0.001
P_t <- matrix(rep(0, 4), nrow=2)
P <- matrix(rep(0, 4), nrow=2)
beta <- matrix(rep(0, nrow(y)*2), ncol=2)
y_fitted <- rep(0, nrow(y))
nu <- rep(0, nrow(y))
Q <- rep(0, nrow(y))
################################################################
# Function to implements the iterative Non-Gaussian filter operations
################################################################
kalman_iteration <- function(y, x) {
for(i in 1:nrow(y)) {
  if(i > 1) {
    beta[i, ] <- beta[i-1, ] # state transition
    P_t <- P + sigma_v # state covariance prediction
    }
  y_fitted[i] <- x[i, ] %*% beta[i, ] # observation prediction
  Q[i] <- x[i, ] %*% P_t %*% t(x[i, ]) + Ve # observation variance prediction
  nu[i] <- y[i] - y_fitted[i] # prediction error
  K_gain <- P_t %*% t(x[i, ]) / Q[i] # information gain
  # updating the state
  beta[i, ] <- beta[i, ] + K_gain * nu[i]
  P <- P_t - K_gain %*% x[i, ] %*% P_t
}
  return(list(beta, P, Q, nu))
}

res <- kalman_iteration(y,x) # Implementation of function
# Extract results
beta <- xts(res[[1]], order.by=index(pair.stock))
plot(beta[2:nrow(beta), 1],type='l',main ='Dynamic hedge ratio',col = "blue")
plot(beta[2:nrow(beta), 2],type='l',main ='Dynamic intercept',col = "blue")
```

```{r}
# plot trade signals
nu <- xts(res[[4]], order.by=index(pair.stock))
sqrtQ <- xts(sqrt(res[[3]]), order.by=index(pair.stock))
signals <- merge(nu, sqrtQ, -sqrtQ)
colnames(signals) <- c("nu", "sqrtQ", "negsqrtQ")
```
### Functions to calculate the optimal value of threshold $p$, which maximizes the Sharpe ratio (SR) and cumulative profit
```{r}
#############################################################
# Required functions to calculate p and profit
#############################################################
# Function to generate positions and calculate profit and loss for multiple stocks
PnL<-function(signals,nu,beta,x,y){
len <- length(index(signals)) 
vec.sig<-ifelse((signals[1:len]$nu > signals[1:len]$sqrtQ) & 
                  (lag.xts(signals$nu, 1) < lag.xts(signals$sqrtQ, 1)), -1, 
         ifelse((signals[1:len]$nu < signals[1:len]$negsqrtQ) & 
                 (lag.xts(signals$nu, 1) > lag.xts(signals$negsqrtQ, 1)), 1, 0))
colnames(vec.sig) <- "vectorise.signals"
# getting only the first signals
vec.sig[vec.sig == 0] <- NA # replace 0 by NA
vec.sig <- na.locf(vec.sig) # replace the missing values by last real observations
vec.sig <- diff(vec.sig)/2
# generate positions and calculate profit for two stocks
if(ncol(beta)==2){
sim <- merge(lag.xts(vec.sig,1), beta[, 1], x[, 1], y)
colnames(sim) <- c("sig", "hedge", assets[1], assets[2])
sim$posX <- sim$sig * -1000 * sim$hedge
sim$posY <- sim$sig * 1000   
sim$posX[sim$posX == 0] <- NA
sim$posX <- na.locf(sim$posX)
sim$posY[sim$posY == 0] <- NA
sim$posY <- na.locf(sim$posY)
PLX <- sim$posX * diff(sim[, assets[1]])
PLY <- sim$posY * diff(sim[, assets[2]])
profit_loss <- PLX + PLY
  }
# generate positions and calculate profit for three stocks
if(ncol(beta)==3){
sim <- merge(lag.xts(vec.sig,1), beta[, 1], beta[, 2], x[, 1], x[, 2], y)
colnames(sim) <- c("sig", "hedge1", "hedge2", assets[1], assets[2], assets[3])
sim$posX1 <- sim$sig * -1000 * sim$hedge1
sim$posX2 <- sim$sig * -1000 * sim$hedge2
sim$posY <- sim$sig * 1000   
sim$posX1[sim$posX1 == 0] <- NA
sim$posX1 <- na.locf(sim$posX1)
sim$posX2[sim$posX2 == 0] <- NA
sim$posX2 <- na.locf(sim$posX2)
sim$posY[sim$posY == 0] <- NA
sim$posY <- na.locf(sim$posY)
PLX <- sim$posX1 * diff(sim[, assets[1]]) + sim$posX2 * diff(sim[, assets[2]])
PLY <- sim$posY * diff(sim[, assets[3]])
profit_loss <- PLX + PLY
   }
# generate positions and calculate profit for four stocks
if(ncol(beta)==4){
sim <- merge(lag.xts(vec.sig,1), beta[,1], beta[,2], beta[,3], x[,1], x[,2], x[,3], y)
colnames(sim) <- c("sig", "hedge1", "hedge2", "hedge3",assets[1],assets[2], 
                   assets[3], assets[4])
sim$posX1 <- sim$sig * -1000 * sim$hedge1
sim$posX2 <- sim$sig * -1000 * sim$hedge2
sim$posX3 <- sim$sig * -1000 * sim$hedge3
sim$posY <- sim$sig * 1000   
sim$posX1[sim$posX1 == 0] <- NA
sim$posX1 <- na.locf(sim$posX1)
sim$posX2[sim$posX2 == 0] <- NA
sim$posX2 <- na.locf(sim$posX2)
sim$posX3[sim$posX3 == 0] <- NA
sim$posX3 <- na.locf(sim$posX3)
sim$posY[sim$posY == 0] <- NA
sim$posY <- na.locf(sim$posY)
PLX <- sim$posX1 * diff(sim[, assets[1]]) + sim$posX2 * diff(sim[, assets[2]]) + 
                              sim$posX3 * diff(sim[, assets[3]])
PLY <- sim$posY * diff(sim[, assets[2]])
profit_loss <- PLX + PLY
    }
return(ProfitLoss=profit_loss)
}
# Functions to calculate the optimal value of threshold, $p$
SR.train<-function(nu, sqrtQ, p){
signals <- merge(nu, p*sqrtQ, -p*sqrtQ)
colnames(signals) <- c("nu", "sqrtQ", "negsqrtQ")
# Implementation of profit
profit.loss<- PnL(signals,nu,beta,x,y)
st_p <- sqrt(252)*mean(na.omit(profit.loss))/sd(na.omit(profit.loss))
return (st_p)
} 
```

```{r}
# determining optimal p to maximize Sharpe ratio 
p <- seq(0.1, 2, 0.01)
SR<-0
for(j in 1:length(p)){
      SR[j] <- SR.train (nu, sqrtQ, p[j])
    }
max(na.omit(SR))
plot(p, SR, type = "l", col = "blue")
```
## Calcualte the optimal value of p by maximizing the SR
```{r}
p.opt<-p[which.max(SR)]
p.opt
```

```{r}
# create optimal trading signals
p <- p.opt
signals <- merge(nu, p*sqrtQ, -p*sqrtQ)
colnames(signals) <- c("nu", "sqrtQ", "negsqrtQ")
plot(signals[3:length(index(signals))], ylab='nu', main = 'Trading signals', 
              col=c('blue', 'red', 'red'), lwd=c(1,2,2))
```
## Calculate the cumulative profit using optimal trading signals
```{r}
# Implementation of profit and loss fucntion to calculate cumulative profit
profit.loss<- PnL(signals,nu,beta,x,y)
sum (na.omit(profit.loss))
plot(cumsum(na.omit(profit.loss)), main="Cumulative profit, $", col = "blue")
```

# Implementation of Non-Gaussian filter for multiple trading (three stocks)

```{r}
assets <- c("EWA", "EWC", "IGE")
three.stocks <- merge(stocks[, 8], stocks[, 9], stocks [, 10])
colnames(three.stocks) <- assets
plot(three.stocks, legend.loc=1)
jotest=ca.jo(three.stocks, type="trace", K=2, ecdet="none", spec="longrun")
summary(jotest)
```
## Implementation of Non-Gaussian filter algorithm
```{r}
x <- three.stocks[, 1:2]
y <- three.stocks[, 3]
x$intercept <- rep(1, nrow(x)) # creating intercept
var_e <- 0.0001 # innovation covariance of observations
sigma_v <- var_e/(1-var_e)*diag(3) # covariance matrix of state
#sigma_v 
Ve <- 0.001
P_t <- matrix(rep(0, 9), nrow=3)
P <- matrix(rep(0, 9), nrow=3)
beta <- matrix(rep(0, nrow(y)*3), ncol=3)
y_fitted <- rep(0, nrow(y))
nu <- rep(0, nrow(y))
Q <- rep(0, nrow(y))
# Implementation of function of iterative Non-Gaussian filter operations
res <- kalman_iteration(y,x)
# Extract results
beta <- xts(res[[1]], order.by=index(three.stocks))
plot(beta[2:nrow(beta), 1:2], type='l', main = 'Dynamic hedge ratios', 
     col = c("blue", "green"))
plot(beta[2:nrow(beta), 3], type='l', main = 'Dynamic updated intercept', col = "blue")
```

```{r}
# plot trade signals
nu <- xts(res[[4]], order.by=index(three.stocks))
sqrtQ <- xts(sqrt(res[[3]]), order.by=index(three.stocks))
signals <- merge(nu, sqrtQ, -sqrtQ)
colnames(signals) <- c("nu", "sqrtQ", "negsqrtQ")
```

```{r}
# determining optimal p to maximize Sharpe ratio 
p <- seq(0.1, 2, 0.01)
SR<-0
for(j in 1:length(p)){
      SR[j] <- SR.train (nu, sqrtQ, p[j])
    }
max(na.omit(SR))
plot(p, SR, type = "l", col = "blue", ylab = "Annualized SR")
```
## Calcualte the optimal value of p by maximizing the SR
```{r}
p.opt<-p[which.max(SR)]
p.opt
```

```{r}
# create optimal trading signals
p <- p.opt
signals <- merge(nu, p*sqrtQ, -p*sqrtQ)
colnames(signals) <- c("nu", "sqrtQ", "negsqrtQ")
plot(signals[3:length(index(signals))], ylab='nu', main = 'Trading signals', 
                col=c('blue', 'red', 'red'), lwd=c(1,2,2))
```
## Calculate the cumulative profit using optimal trading signals
```{r}
# Implementation of profit and loss fucntion to calculate cumulative profit
profit.loss<- PnL(signals,nu,beta,x,y)
sum (na.omit(profit.loss))
plot(cumsum(na.omit(profit.loss)), main="Cumulative profit, $", col = "blue")
```

# Implementation of Non-Gaussian filter for multiple trading (four stocks)

## Plot and test for the presence of cointegration between stocks 
```{r}
assets <- c("GDX", "EWA", "EWC", "IGE")
four.stocks <- merge(stocks[, 7], stocks[, 8], stocks[, 9], stocks [, 10])
colnames(four.stocks) <- assets
plot(four.stocks, legend.loc=1)
jotest=ca.jo(four.stocks, type="trace", K=2, ecdet="none", spec="longrun")
summary(jotest)
```
## Implementation of Non-Gaussian filter algorithm
```{r}
x <- four.stocks[, 1:3]
y <- four.stocks[, 4]
x$intercept <- rep(1, nrow(x)) # Creating intercept
var_e <- 0.0001 # innovation covariance of observations
sigma_v <- var_e/(1-var_e)*diag(4) # covariance matrix of state
#sigma_v 
Ve <- 0.001
P_t <- matrix(rep(0, 16), nrow=4)
P <- matrix(rep(0, 16), nrow=4)
beta <- matrix(rep(0, nrow(y)*4), ncol=4)
y_fitted <- rep(0, nrow(y))
nu <- rep(0, nrow(y))
Q <- rep(0, nrow(y))
# Implementation of function of iterative Non-Gaussian filter operations
res <- kalman_iteration(y,x)
# Extract results
beta <- xts(res[[1]], order.by=index(four.stocks))
plot(beta[2:nrow(beta), 1:3], type='l', main = 'Dynamic updated hedge ratios', 
     col = c("blue", "green", "red"))
plot(beta[2:nrow(beta), 4], type='l', main = 'Dynamic updated intercept', col = "blue")
```

```{r}
# plot trading signals
nu <- xts(res[[4]], order.by=index(four.stocks))
sqrtQ <- xts(sqrt(res[[3]]), order.by=index(four.stocks))
signals <- merge(nu, sqrtQ, -sqrtQ)
colnames(signals) <- c("nu", "sqrtQ", "negsqrtQ")
```

```{r}
# determining optimal p to maximize Sharpe ratio 
p <- seq(0.1, 2, 0.01)
SR<-0
for(j in 1:length(p)){
      SR[j] <- SR.train (nu, sqrtQ, p[j])
    }
max(na.omit(SR))
plot(p, SR, type = "l", col = "blue", ylab = "Annualized SR")
```
## Calcualte the optimal value of p by maximizing the SR
```{r}
p.opt<-p[which.max(SR)]
p.opt
```

```{r}
# create optimal trading signals
p <- p.opt
signals <- merge(nu, p*sqrtQ, -p*sqrtQ)
colnames(signals) <- c("nu", "sqrtQ", "negsqrtQ")
plot(signals[3:length(index(signals))], ylab='nu', main = 'Trading signals', 
                col=c('blue', 'red', 'red'), lwd=c(1,2,2))
```
## Calculate the cumulative profit using optimal trading signals
```{r}
# Implementation of profit and loss fucntion to calculate cumulative profit
profit.loss<- PnL(signals,nu,beta,x,y)

plot(cumsum(na.omit(profit.loss)), main="Cumulative profit, $", col = "blue")
sum (na.omit(profit.loss))
```
# Buy and hold strategy
```{r}
maxx<- nrow (stocks)
GDX <- as.numeric(stocks [maxx, 7]) - as.numeric(stocks [1, 7])
EWA <- as.numeric(stocks [maxx, 8]) - as.numeric(stocks [1, 8])
EWC <- as.numeric(stocks [maxx, 9]) - as.numeric(stocks [1, 9])
IGE <- as.numeric(stocks [maxx, 10]) - as.numeric(stocks [1, 10])
1000*(EWA + EWC)
1000*(EWA + EWC + IGE)
1000*(GDX + EWA + EWC + IGE)
```
